Semantic Linearity in a Lazy Optimising Compiler

### 2025-02-26 icfp.tex
appendix shuffle

The Launchbury-style natural semantics, presented in Figure~\ref{fig:opsem}, captures
standard lazy evaluation while ignoring linearity information.
% The second
% semantics is significantly instrumented with type information and is thus
% \emph{linearity aware}, resulting in a stuck state when a linear
% variable is used more than once. The former is a more direct model of
% an implementation of lazy evaluation, however it is not so convenient
% to reason about formally. The latter semantics is augmented with the additional
% information necessary to strengthen the induction hypothesis to make our
% type safety proofs go through.
% %
% We show that for well-typed terms, the two semantics coincide. 
%
%Both semantics are
The natural semantics is equipped with an evaluation environment that
maps variables to expressions and is
mutated in order to express evaluation\todo{aqui não seria sharing?}. Terms are explicit about
their sharing (in the sense of shared reductions in lazy evaluation),
and so we perform a translation step that makes all sharing explicit
through let-binders (see Appendix~\ref{app:opsem}).
We write $\llet{\ov{x_i{:}_{\Delta_i}\sigma = e_i}}{e'}$ for the iterated
let-binding of variables $\ov{x_i}$.

@@

The instrumented semantics is presented in
Figure~\ref{fig:linopsem}, defining the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
\mid v) : \sigma , \Sigma$ where $\Gamma$ and $\Delta$ are typing
contexts for expressions in $\Sigma$
($\Gamma$ tracks unrestricted and $\Delta$-bound variables, $\Delta$ tracks linear
assumptions);
$\Theta$ and $\Theta'$ are evaluation environments, consisting of bindings of the form
$x :_\pi \sigma = e$ or $x :_\Delta \sigma = e$; $e$ is the expression being evaluated and $v$
its resulting value, both of type $\sigma$; $\Sigma$ is a list of assignments of the form
$e :_\pi \tau$ or $x :_\Delta \tau$\todo{isto não me fez sentido} which are expressions in which
bindings in $\Theta$ may also be used. This additional data is needed to
inductively show that the overall evaluation state is well-typed (and
linearity preserving). The rules mostly mirror those of the natural
semantics, with the exception of the linear variable rule which erases
the binding from the execution environment. Both semantics define
evaluation relations of the form $a \Downarrow b$, where $a$ and $b$
are evaluation states. We also rely on a notion
of partial derivations and partial evaluation (see
Appendix~\ref{app:opsem}), written $a\Downarrow^* b$, which allows us
to state and reason about a progress property.\todo{Na semantica O omega e 1 ainda estão separados, é assim?}

@@

       \infer[]
    {\Gamma ; \Delta  \vdash (\Theta  , x:_{\Delta'} \sigma = e\mid e)
    \Downarrow (\Theta' \mid v) : \sigma , \Sigma}
    {\Gamma ; \Delta \vdash (\Theta , x:_{\Delta'} \sigma = e \mid x)
      \Downarrow (\Theta' , x:_{\Delta''} \sigma = v \mid v) : \sigma, \Sigma}
    \\[1.5em]
           \infer[]
    {\Gamma ;\Delta  \vdash (\Theta \mid e)
    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
    {\Gamma ; \Delta \vdash (\Theta , x:_1 \sigma = e \mid x)
    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
    \\[1.5em]

@@

    \infer[]
    {K : \ov{ \sigma_i
    \rightarrow_\omega \sigma_j \lolli} T\\
    \Gamma ; \Delta, z{:}_1\s \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
    K~\ov{x_i}) : T , e', \Sigma \quad
    \Gamma ;\Delta \vdash (\Theta' \mid  e'[K~\ov{x_i}/z])
    \Downarrow (\Theta'' \mid v) : \tau,\Sigma }
    {\Gamma ; \Delta \vdash (\Theta \mid 
    \ccase{e}{z{:}_{\Delta'}T~\{\dots , \_\to
    e'}\}) \Downarrow (\Theta'' \mid v) : \tau}
    \\[1.5em]
    \infer[]
    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % \Downarrow \Theta_n : v_n \quad
    \Gamma ; \Delta \vdash (\Theta, \overline{x_i:_\omega \sigma_i =
    e_i} \mid e' )\Downarrow (\Theta'
    \mid v) : \tau
    }
    {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'})
    \Downarrow (\Theta' \mid v) : \tau}

@@

%For example, we might float an expensive computation involving $z$ out of the
%case alternative, where $z$ is now out of scope:
%
%{\small\[
%\begin{array}{l}
%(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to \ E[expensive~z] \} }}{\dots})
%\Longrightarrow_\textrm{Reverse binder swap}\\
%(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[expensive~x] \} }}{\dots})
%\Longrightarrow_\textrm{Commuting case-let}\\
%\lambda x.~\llet{t = \mathit{expensive}~x}{(\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[t] \} }}{\dots})}\\
%\end{array}
%\]}
%
%\noindent If $go$ is compiled into a loop, we only compute
%$\mathit{expensive}~x$ once instead of in every loop iteration.
%
Even though GHC applies the reverse binder swap transformation in the
Core-to-Core passes, this optimisation violates linearity in Linear Core
because the transformed program is rejected -- the scrutinee $x$, a single
variable, is not considered to be in WHNF by our system, thus occurrences of
$x$ in any alternative are ill-typed in it. This is a conservative choice in $\lambda^\pi_\Delta$ for
typing the subtle case where the scrutinee is a single variable:

Consider $\ccase{x}{\_ \to x}$. Perhaps surprisingly, this expression can be
considered semantically linear in a call-by-need calculus but violates
linearity in a call-by-name one:
%
%An expression in which a variable $x$ occurs both in the case scrutinee and in
%the alternatives (e.g.,~$\ccase{x}{\_ \to x}$) could be considered linear if
%considered in isolation:
% \[
% \G; \xl \vdash \ccase{x}{\_ \to x} : \s
% \]
% The reasoning is done by branching on whether $x$ refers to an expression in
% WHNF or an unevaluated thunk.
% \begin{itemize}
% \item
Under call-by-need, if $x$ refers to an unevaluated expression, scrutinizing it forces the expression
bound by $x$ to WHNF. In a subsequent use of
$x$ in an alternative, $x$ refers to the already evaluated scrutinee in WHNF.
Since $x$ is just another name (like the case binder) for the scrutinee in
WHNF, its use is also valid in the alternatives.
%
Dually, if $x$ refers to an expression already in WHNF, no evaluation takes
place and $x$ is still just another name for the scrutinee in WHNF.

@@

@@

%\input{language-v4/proofs/TypePreservationTheorem}

@@

\section{Instrumented Operational Semantics}\label{app:opsem}

In this section we develop a lazy big-step operational semantics for $\lambda_\Delta^\pi$ in the
style of Launchbury~\cite{10.1145/158511.158618}. The semantics is
equipped with an environment which assigns values to variables and is
mutated in order to express lazy evaluation. Terms are explicit about
their sharing (in the sense of shared reductions in lazy evaluation),
and so we perform a translation step that makes all sharing explicit
through let-binders.
Below we write $\llet{\ov{x_i{:}_{\Delta_i}\sigma = e_i}}{e'}$ for the iterated
let-binding of variables $\ov{x_i}$.
\[
  \begin{array}{lclr}
    x^* & \triangleq & x\\

    (\Lambda p.~e)^* & \triangleq & \Lambda p.~(e)^*\\
    (\lambda x:_\pi\sigma .e)^* & \triangleq & \lambda x:_\pi\sigma . (e)^*\\
    (e~\pi)^* & \triangleq & (e)^*~\pi\\
    (e~x)^* & \triangleq & (e)^*~x\\
    (e~e')^* & \triangleq & \llet{x{:}_{\Delta}\sigma =
                            (e')^*}{(e)^*~x}     \\ &&\qquad \mbox{with
                            $\Gamma;\Delta' \vdash e :
                            \sigma\rightarrow_\pi \varphi$} \\ &&\qquad \mbox{and
                                       $\Gamma;\Delta \vdash e' : \sigma$}\\
      K~\ov{e_i} & \triangleq & \llet{\ov{x_i:_{\Delta_i} \sigma_i =
                                   (e_i)^*}}{K~\ov{x_i}} \\ &&
                                       \qquad\mbox{with
                                       $\ov{\Gamma;\Delta_i\vdash
                                       e_i :
                                       \sigma_i}$}\\                                              
(\ccase{e}{z{:}_{\Delta}\sigma~\{\overline{\rho\to e'}\}})^* &
                                                               \triangleq
                                &
                                  \ccase{(e)^*}{z{:}_{\Delta}\sigma~\{\overline{\rho\to (e')^*}\}}\\
    (\llet{x{:}_{\Delta}\sigma = e}{e'}      )^* & \triangleq & \llet{x{:}_{\Delta}\sigma  = (e)^*}{(e')^*}    \\
    (\lletrec{\overline{x{:}_{\Delta}\sigma = e}}{e'}  )^* & \triangleq &\lletrec{\overline{x{:}_{\Delta}\sigma = (e)^*}}{(e')^*}  
      \end{array}
\]

The big-step operational semantics is defined via the relation
$\Theta : e \Downarrow \Theta' : v$ where $e$ is an expression, $v$ a
value, $\Theta$ and $\Theta'$ are environments with bindings of the
form $x:_\omega\sigma = e$ which assigns the expression $e$ to the
variable $x$ of the given type.

\[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Theta : \Lambda p.e \Downarrow \Theta : \Lambda p.e}
    \qquad
    \infer[]
    {\Theta  : e\Downarrow \Theta' : \Lambda p.e' \quad \Theta' :
    e'[\pi/p]\Downarrow \Theta'' : v}
    {\Theta : e~\pi\Downarrow \Theta'' : v }\\[1.5em]
    \infer[]
    {\,}
    {\Theta : \lambda x:_\pi \sigma . e \Downarrow \Theta : \lambda x:_\pi \sigma . e }
    \qquad
    \infer[]
    {\Theta : e \Downarrow \Theta' : \lambda y:_\pi \sigma . e' \quad
    \Theta' : e'[x/y] \Downarrow \Theta'' : v }
    {\Theta : e~x \Downarrow \Theta'' : v}\\[1.5em]
    \infer[]
    {(\Theta , x:_\omega \sigma = e): e \Downarrow \Theta' : v}
    {(\Theta , x:_\omega \sigma = e) : x \Downarrow (\Theta',x:_\omega
    \sigma = v) : v}
    \qquad
    \infer[]
    {(\Theta,x:_\omega \sigma = e) : e' \Downarrow \Theta' : v}
    {\Theta : \llet{x{:}_{\Delta}\sigma = e}{e'}\Downarrow \Theta' : v
    }\\[1.5em]
    \infer[]
    {\,}
    {\Theta : K~\ov{x_i} \Downarrow \Theta : K~\ov{x_i}}
    \qquad
    \infer[]
    {\Theta : e \Downarrow \Theta' : K~\ov{x_i} \quad
    \Theta' : e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] \Downarrow \Theta'' : v }
    {\Theta :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i} \to e', {\dots}}\}}
    \Downarrow \Theta'' : v}\\[1.5em]
    \infer[]
    {\Theta : e \Downarrow \Theta' : K~\ov{x_i} \quad
    \Theta' : e'[K~\ov{x_i}/z] \Downarrow \Theta'' : v }
    {\Theta :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
    e'}\} \Downarrow \Theta'' : v}
    \qquad
    \infer[]
    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % \Downarrow \Theta_n : v_n \quad
    (\Theta, \overline{x_i:_\omega \sigma_i = e_i}) : e' \Downarrow \Theta'
    : v
    }
    {\Theta : \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'}
    \Downarrow \Theta' : v}
    \end{array}
\]

We follow an approach
similar to that of \cite{cite:linearhaskell}, defining a linearity
aware operational semantics in which linear variables are erased from
the runtime environment, ensuring that any term violating linearity
will result in a stuck state. The semantics is type-aware, defined by
the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
\mid v) : \sigma , \Sigma$ where $\Gamma$ and $\Delta$ are typing judgments
($\Gamma$ tracks unrestricted variables, $\Delta$ tracks linear
assumptions);
$\Theta$ and $\Theta'$ are runtime environments, consisting of bindings of the form
$x :_\pi \sigma = e$; $e$ is the expression being evaluated and $v$
its resulting value, both of type $\sigma$; $\Sigma$ is a list of assignments of the form
$e :_\pi \tau$ or $x :_\Delta \tau$ which are expressions in which variables in $\Theta$ are used.

>>>>

The Launchbury-style natural semantics, presented in Figure~\ref{fig:opsem}, captures
standard lazy evaluation while ignoring linearity information.
% The second
% semantics is significantly instrumented with type information and is thus
% \emph{linearity aware}, resulting in a stuck state when a linear
% variable is used more than once. The former is a more direct model of
% an implementation of lazy evaluation, however it is not so convenient
% to reason about formally. The latter semantics is augmented with the additional
% information necessary to strengthen the induction hypothesis to make our
% type safety proofs go through.
% %
% We show that for well-typed terms, the two semantics coincide. 
%
%Both semantics are
The natural semantics is equipped with an evaluation environment that
maps variables to expressions and is
mutated in order to express shared evaluation. Terms are explicit about
their sharing (in the sense of shared reductions in lazy evaluation),
and so we perform a translation step that makes all sharing explicit
through let-binders (see Appendix~\ref{app:opsem}).
We write $\llet{\ov{x_i{:}_{\Delta_i}\sigma = e_i}}{e'}$ for the iterated
let-binding of variables $\ov{x_i}$.

@@

The instrumented semantics is presented in
Figure~\ref{fig:linopsem}, defining the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
\mid v) : \sigma , \Sigma$ where $\Gamma$ and $\Delta$ are typing
contexts for expressions in $\Sigma$
($\Gamma$ tracks unrestricted and $\Delta$-bound variables, $\Delta$ tracks linear
assumptions);
$\Theta$ and $\Theta'$ are evaluation environments, consisting of bindings of the form
$x :_\pi \sigma = e$ or $x :_\Delta \sigma = e$; $e$ is the expression being evaluated and $v$
its resulting value, both of type $\sigma$; $\Sigma$ is a list of assignments of the form
$e : \tau$ which are expressions in which
bindings in $\Theta$ may also be used. This additional data is needed to
inductively show that the overall evaluation state is well-typed (and
linearity preserving). The rules mostly mirror those of the natural
semantics, with the exception of the linear variable rule which erases
the binding from the execution environment. Both semantics define
evaluation relations of the form $a \Downarrow b$, where $a$ and $b$
are evaluation states. We also rely on a notion
of partial derivations and partial evaluation (see
Appendix~\ref{app:opsem}), written $a\Downarrow^* b$, which allows us
to state and reason about a progress property.\todo{Na semantica O omega e 1 ainda estão separados, é assim?}

@@

       \infer[]
    {\Gamma ; \Delta  \vdash (\Theta  , x:_{\Delta'} \sigma = e\mid e)
    \Downarrow (\Theta' \mid v) : \sigma , \Sigma \and \D'' = \D' \setminus (\Theta \setminus \Theta')}
    {\Gamma ; \Delta \vdash (\Theta , x:_{\Delta'} \sigma = e \mid x)
      \Downarrow (\Theta' , x:_{\Delta''} \sigma = v \mid v) : \sigma,
    \Sigma }
    \\[1.5em]
           \infer[]
    {\Gamma ;\Delta  \vdash (\Theta \mid e)
    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
    {\Gamma ; \Delta \vdash (\Theta , x:_1 \sigma = e \mid x)
    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
    \\[1.5em]

@@

    \infer[]
    {K : \ov{ \sigma_i
    \rightarrow_\omega \sigma_j \lolli} T\\
    \Gamma ; \Delta, z{:}_1\s \vdash (\Theta \mid e) \Downarrow (\Theta' \mid
    K~\ov{x_i}) : T , e', \Sigma \quad
    \Gamma ;\Delta \vdash (\Theta' \mid  e'[K~\ov{x_i}/z])
    \Downarrow (\Theta'' \mid v) : \tau,\Sigma }
    {\Gamma ; \Delta \vdash (\Theta \mid 
    \ccase{e}{z{:}_{\Delta'}T~\{\dots , \_\to
    e'}\}) \Downarrow (\Theta'' \mid v) : \tau}
    \\[1.5em]
    \infer[]
    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % \Downarrow \Theta_n : v_n \quad
    \Gamma ; \Delta \vdash (\Theta, \overline{x_i:_{\Delta'} \sigma_i =
    e_i} \mid e' )\Downarrow (\Theta'
    \mid v) : \tau
    }
    {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta'}\sigma_i = e_i}}{e'})
    \Downarrow (\Theta' \mid v) : \tau}

@@

%For example, we might float an expensive computation involving $z$ out of the
%case alternative, where $z$ is now out of scope:
%
%{\small\[
%\begin{array}{l}
%(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to \ E[expensive~z] \} }}{\dots})
%\Longrightarrow_\textrm{Reverse binder swap}\\
%(\lambda x.~\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[expensive~x] \} }}{\dots})
%\Longrightarrow_\textrm{Commuting case-let}\\
%\lambda x.~\llet{t = \mathit{expensive}~x}{(\lletrec{go~y = \ccase{x}{z~\{(a,b) \to E[t] \} }}{\dots})}\\
%\end{array}
%\]}
%
%\noindent If $go$ is compiled into a loop, we only compute
%$\mathit{expensive}~x$ once instead of in every loop iteration.
%
Even though GHC applies the reverse binder swap transformation in the
Core-to-Core passes, this optimisation violates linearity in Linear Core
because the transformed program is rejected -- the scrutinee $x$, a single
variable, is not considered to be in WHNF by our system, thus occurrences of
$x$ in any alternative are ill-typed in it. This is a conservative choice in $\lambda^\pi_\Delta$ for
typing the subtle case where the scrutinee is a single variable.

Consider $\ccase{x}{\_ \to x}$. Perhaps surprisingly, this expression can be
considered semantically linear under call-by-need but violates
linearity under call-by-name:
%
%An expression in which a variable $x$ occurs both in the case scrutinee and in
%the alternatives (e.g.,~$\ccase{x}{\_ \to x}$) could be considered linear if
%considered in isolation:
% \[
% \G; \xl \vdash \ccase{x}{\_ \to x} : \s
% \]
% The reasoning is done by branching on whether $x$ refers to an expression in
% WHNF or an unevaluated thunk.
% \begin{itemize}
% \item
Under call-by-need, if $x$ refers to an unevaluated expression, scrutinizing it forces the expression
bound by $x$ to WHNF. In a subsequent use of
$x$ in an alternative, $x$ refers to the already evaluated scrutinee in WHNF.
Since $x$ is just another name (like the case binder) for the scrutinee in
WHNF, its use is also valid in the alternatives.
%
Dually, if $x$ refers to an expression already in WHNF, no evaluation takes
place and $x$ is still just another name for the scrutinee in WHNF.

@@

\section{Natural and Instrumented Operational Semantics}\label{app:opsem}

We define a lazy big-step operational semantics for $\lambda_\Delta^\pi$ in the
style of Launchbury~\cite{10.1145/158511.158618} natural semantics. The semantics is
equipped with an environment which assigns expressions to variables and is
mutated in order to express lazy evaluation. Terms are explicit about
their sharing (in the sense of shared reductions in lazy evaluation),
and so we perform a translation that makes all sharing explicit
through let-binders. This translation is presented in Figure~\ref{fig:explicitsharing}.
We write $\llet{\ov{x_i{:}_{\Delta_i}\sigma = e_i}}{e'}$ for the iterated
let-binding of variables $\ov{x_i}$.
\begin{figure}
\[
  \begin{array}{lclr}
    x^* & \triangleq & x\\

    (\Lambda p.~e)^* & \triangleq & \Lambda p.~(e)^*\\
    (\lambda x:_\pi\sigma .e)^* & \triangleq & \lambda x:_\pi\sigma . (e)^*\\
    (e~\pi)^* & \triangleq & (e)^*~\pi\\
    (e~x)^* & \triangleq & (e)^*~x\\
    (e~e')^* & \triangleq & \llet{x{:}_{\Delta}\sigma =
                            (e')^*}{(e)^*~x}     \\ &&\qquad \mbox{with
                            $\Gamma;\Delta' \vdash e :
                            \sigma\rightarrow_\pi \varphi$} \\ &&\qquad \mbox{and
                                       $\Gamma;\Delta \vdash e' : \sigma$}\\
      K~\ov{e_i} & \triangleq & \llet{\ov{x_i:_{\Delta_i} \sigma_i =
                                   (e_i)^*}}{K~\ov{x_i}} \\ &&
                                       \qquad\mbox{with
                                       $\ov{\Gamma;\Delta_i\vdash
                                       e_i :
                                       \sigma_i}$}\\                                              
(\ccase{e}{z{:}_{\Delta}\sigma~\{\overline{\rho\to e'}\}})^* &
                                                               \triangleq
                                &
                                  \ccase{(e)^*}{z{:}_{\Delta}\sigma~\{\overline{\rho\to (e')^*}\}}\\
    (\llet{x{:}_{\Delta}\sigma = e}{e'}      )^* & \triangleq & \llet{x{:}_{\Delta}\sigma  = (e)^*}{(e')^*}    \\
    (\lletrec{\overline{x{:}_{\Delta}\sigma = e}}{e'}  )^* & \triangleq &\lletrec{\overline{x{:}_{\Delta}\sigma = (e)^*}}{(e')^*}  
      \end{array}
    \]
    \caption{Explicit Sharing Translation\label{fig:explicitsharing}}
  \end{figure}

The natural semantics (Figure~\ref{fig:natsemapp}) is defined via the relation
$\Theta : e \Downarrow \Theta' : v$ where $e$ is an expression, $v$ a
value, $\Theta$ and $\Theta'$ are environments with bindings of the
form $x:_\omega\sigma = e$ which assigns the expression $e$ to the
variable $x$ of the given type.

\begin{figure}
\[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Theta : \Lambda p.e \Downarrow \Theta : \Lambda p.e}
    \qquad
    \infer[]
    {\Theta  : e\Downarrow \Theta' : \Lambda p.e' \quad \Theta' :
    e'[\pi/p]\Downarrow \Theta'' : v}
    {\Theta : e~\pi\Downarrow \Theta'' : v }\\[1.5em]
    \infer[]
    {\,}
    {\Theta : \lambda x:_\pi \sigma . e \Downarrow \Theta : \lambda x:_\pi \sigma . e }
    \qquad
    \infer[]
    {\Theta : e \Downarrow \Theta' : \lambda y:_\pi \sigma . e' \quad
    \Theta' : e'[x/y] \Downarrow \Theta'' : v }
    {\Theta : e~x \Downarrow \Theta'' : v}\\[1.5em]
    \infer[]
    {(\Theta , x:_\omega \sigma = e): e \Downarrow \Theta' : v}
    {(\Theta , x:_\omega \sigma = e) : x \Downarrow (\Theta',x:_\omega
    \sigma = v) : v}
    \qquad
    \infer[]
    {(\Theta,x:_\omega \sigma = e) : e' \Downarrow \Theta' : v}
    {\Theta : \llet{x{:}_{\Delta}\sigma = e}{e'}\Downarrow \Theta' : v
    }\\[1.5em]
    \infer[]
    {\,}
    {\Theta : K~\ov{x_i} \Downarrow \Theta : K~\ov{x_i}}
    \qquad
    \infer[]
    {\Theta : e \Downarrow \Theta' : K~\ov{x_i} \quad
    \Theta' : e'\ov{[x_i/y_i]}[K~\ov{x_i}/z] \Downarrow \Theta'' : v }
    {\Theta :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{{\dots,K~\ov{y_i} \to e', {\dots}}\}}
    \Downarrow \Theta'' : v}\\[1.5em]
    \infer[]
    {\Theta : e \Downarrow \Theta' : K~\ov{x_i} \quad
    \Theta' : e'[K~\ov{x_i}/z] \Downarrow \Theta'' : v }
    {\Theta :
    \ccase{e}{z{:}_{\Delta'}\sigma~\{\dots , \_\to
    e'}\} \Downarrow \Theta'' : v}
    \qquad
    \infer[]
    {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
    % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
    % \Downarrow \Theta_n : v_n \quad
    (\Theta, \overline{x_i:_\omega \sigma_i = e_i}) : e' \Downarrow \Theta'
    : v
    }
    {\Theta : \lletrec{\overline{x_i{:}_{\Delta}\sigma_i = e_i}}{e'}
    \Downarrow \Theta' : v}
    \end{array}
  \]
  \caption{Natural Semantics for $\lambda_\Delta^\pi$\label{fig:natsemapp}}
\end{figure}

Our instrumented, linearity-aware semantics (akin to that
of~\cite{cite:linearhaskell})
is defined in Figure~\ref{fig:instrsemapp}.
In this semantics, linear variables are erased from
the runtime environment, ensuring that any term violating linearity
produces a stuck state. The semantics is defined by
the judgment $\Gamma ; \Delta \vdash (\Theta \mid e) \Downarrow (\Theta'
\mid v) : \sigma , \Sigma$ where $\Gamma$ and $\Delta$ are typing judgments
($\Gamma$ tracks unrestricted variables, $\Delta$ tracks linear
assumptions) for variables occurring in $\Sigma$;
$\Theta$ and $\Theta'$ are runtime environments, consisting of bindings of the form
$x :_\pi \sigma = e$ or $x :_\Delta \sigma = e$ ; $e$ is the expression being evaluated and $v$
its resulting value, both of type $\sigma$; $\Sigma$ is a list of assignments of the form
$e : \tau$, which are expressions in which variables in $\Theta$ are
used.

\begin{figure}
\[
  \begin{array}{c}
    \infer[]
    {\,}
    {\Gamma;\Delta \vdash (\Theta \mid \Lambda p.e) \Downarrow (\Theta \mid
    \Lambda p.e) : \forall p . \sigma , \Sigma }
    \\[1.5em]
    \infer[]
    {\Gamma ; \Delta \vdash (\Theta \mid  e) \Downarrow (\Theta' \mid
    \Lambda p.e') : \forall p . \sigma , \Sigma \quad \Gamma ;\Delta \vdash
      (\Theta' \mid e'[\pi/p]) \Downarrow (\Theta'' \mid v) : \sigma[\pi/p], \Sigma}
    {\Gamma ; \Delta \vdash (\Theta \mid e~\pi) \Downarrow (\Theta''
      \mid v) : \sigma[\pi/p] , \Sigma  }\\[1.5em]
    \infer[]
    {\,}
    {\Gamma ;\Delta \vdash (\Theta \mid \lambda x:_\pi \sigma . e)
    \Downarrow (\Theta \mid \lambda x:_\pi\sigma .e) : \sigma
    \rightarrow_\pi \tau,\Sigma }\\[1.5em]
    \infer[]
    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
    \lambda y:_1 \sigma . e') : \sigma \rightarrow_1 \tau , x:\sigma, \Sigma
    \quad
    \Gamma ; \Delta \vdash (\Theta' , y:_{1} \sigma = x \mid e') \Downarrow (\Theta''
    \mid v) : \tau , \Sigma  }
    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta''
    \mid v) : \tau , \Sigma }\\[1.5em]
        \infer[]
    {\Gamma;\Delta \vdash  (\Theta \mid e) \Downarrow (\Theta' \mid
    \lambda y:_\omega \sigma . e') : \sigma \rightarrow_\omega \tau , x:\sigma, \Sigma
    \quad
            \Gamma ; \Delta \vdash (\Theta' \mid e'[x/y]) \Downarrow (\Theta''
    \mid v) : \tau , \Sigma  }
    {\Gamma ; \Delta\vdash (\Theta \mid  e~x ) \Downarrow (\Theta''
    \mid v) : \tau , \Sigma }\\[1.5em]
       \infer[]
    {\Gamma ; \Delta  \vdash (\Theta, x:_{\Delta'} \sigma = e \mid e)
    \Downarrow (\Theta' \mid v) : \sigma , \Sigma \\
      \D'' = \D' \setminus (\Theta \setminus \Theta')
      }
    {\Gamma ; \Delta \vdash (\Theta , x:_{\Delta'} \sigma = e \mid x)
      \Downarrow (\Theta' , x:_{\Delta''} \sigma = v \mid v) : \sigma, \Sigma}
    \\[1.5em]
           \infer[]
    {\Gamma ;\Delta  \vdash (\Theta \mid e)
    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
    {\Gamma ; \Delta \vdash (\Theta , x:_1 \sigma = e \mid x)
    \Downarrow (\Theta' \mid v) : \sigma,\Sigma}
    \\[1.5em]

     \infer[]
     {% (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_1 \Downarrow \Theta_1
     % : v_1 \dots  (\Gamma, \overline{x_i:_\omega \sigma_i = e_i}) : e_n
     % \Downarrow \Theta_n : v_n \quad
     (\Gamma ; \Delta \vdash (\Theta, \overline{x_i:_{\Delta'} \sigma_i =
     e_i} \mid e' )\Downarrow (\Theta'
     \mid v) : \tau
     }
     {\Gamma ;\Delta \vdash (\Theta \mid \lletrec{\overline{x_i{:}_{\Delta'}\sigma_i = e_i}}{e'})
     \Downarrow (\Theta' \mid v) : \tau}

   \end{array}
 \]
 \caption{Instrumented Semantics for $\lambda_\Delta^\pi$\label{fig:instrsemapp}}
 \end{figure}

Below, we write $\Theta {\uparrow^1}$ for the linear bindings in evaluation environment $\Theta$ and $\Theta {\uparrow^\omega}$ for all other bindings.

We note that for the treatment of recursive let groups, environment
expansion needs to happen for the entire group of let-bound
variables. This is easily achieved by extending the runtime
environment with information about recursive let groupings, which we
omit for the sake of presentation.

\subsection{Bisimulation}

\begin{definition}[Bisimulation]
We define the relation $\gamma(\Theta : e)(\G ; \D \vdash \Theta' \mid
e' : \tau,\Sigma)$ between a well-typed state and an ordinary state of
the lazy big-step semantics if $e = e'$ and $\Theta'{\uparrow^\omega}
\leq \Theta$.
\end{definition}

@@

% \input{language-v4/proofs/TypePreservationTheorem}
\begin{theorem}[Type Preservation]\label{thm:typepres}
  Let $\G ; \D \vdash (\Theta \mid e) \Downarrow (\Theta' \mid v) :
  \tau , \Sigma$ or $\G ; \D \vdash (\Theta \mid e) \Downarrow^* (\Theta' \mid v) :
  \tau , \Sigma$.
  If $\G ; \D \vdash \Theta \mid e : \tau , \Sigma$ then $\G ; \D \vdash \Theta' \mid v : \tau , \Sigma$

We also derive type preservation for the lazy natural semantics.

@@
